---
title: "Homework 2"
output: github_document
---
_Note: This homework uses the `tidyverse` and `dplyr` libraries._
```{r setup, include = FALSE}
library(tidyverse)
library(dplyr)
```

## Problem 1

```{r data, include = FALSE}
nyctransit_df = read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
nyctransit_df #1,868 entries and 32 columns.
```

The `nyctransit_df` dataset (`r ncol(nyctransit_df)` columns x `r nrow(nyctransit_df)` rows) was created based on the _"NYC_Transit_Subway_Entrance_And_Exit_Data.csv"_ file.

```{r summary, include = FALSE}
summary(nyctransit_df)
nyctransit_df = janitor::clean_names(nyctransit_df) #transform variable names to snake case
```

After reading the `nyctransit_df` dataset and taking a look at its 32 variables with `summary(nyctransit_df)`, I applied the snake case convention to the variable names using `clean_names` from the `janitor` package. 

```{r select, include = FALSE}
nyctransit_q1 = 
  select(nyctransit_df, line:entry, vending, ada) |>
   mutate(
     entry = ifelse(entry == "YES", TRUE, FALSE)
     ) #convert entry to logical variable

nyctransit_q1 #1,868 entries and 19 columns. 11 character, 2 logical and 6 double variables
summary(nyctransit_q1)
```

Using the `select()` function, I selected my variables of interest and converted the `entry` variable in this dataset from character (YES vs NO) to a logical variable (TRUE vs. FALSE). 

The resulting `nyctransit_q1` dataset has 19 columns and 1,868 rows. The 19 variables include 11 characters, 2 logicals, and 6 doubles. 
